Change in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-11-25 analysis/paper/paper.Rmd
reorg data files, make & save the map

Here is a citation

@@

Theory and background to our three questions
New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region. Previous studies show that new technologies play an important role in the change in the use of the landscape. For instance, Morisaki et al. argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015]. In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994]. New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group. The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011]. 
The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].

@@

Regional Archaeology

Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia
The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic. In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois [@gao2002critique; @Bar-Yosef_2002; @mellars2007rethinking; @norton2009evolution; @norton2009east; @lee2013current]]. Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].

@@

```{r}
#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
#This is what I made first before you made the recent one

korean_archaeological_sites <- readxl::read_excel("~/koreapaleolithicmobilityoccupation/korean-archaeologica-sites.xlsx")

@@

@@

Statistical methods 
To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type. The elevation is the main proxy for environmental information in this research. The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [####Han 2008####; @Chang_2013]. Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ [###Han 2008###; ####Seong 2008####; ###Choi 2011####; @Chang_2013; @im2015study; @bak2017late]. In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula. In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017]. In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk. This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.

>>>>

@@

## Theory and background to our three questions

New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region. Previous studies show that new technologies play an important role in the change in the use of the landscape. For instance, Morisaki et al. argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015]. In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994]. New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group. The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011].

The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].

@@

## Regional Archaeology

Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia

The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic. In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois [@gao2002critique; @Bar-Yosef_2002; @mellars2007rethinking; @norton2009evolution; @norton2009east; @lee2013current]]. Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].

@@

```{r cache = TRUE}
#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
#This is what I made first before you made the recent one

library(here)

korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))

@@

ggsave(here("analysis/figures/site-map.png"))

@@

Statistical methods

To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type. The elevation is the main proxy for environmental information in this research. The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [####Han 2008####; @Chang_2013]. Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ [###Han 2008###; ####Seong 2008####; ###Choi 2011####; @Chang_2013; @im2015study; @bak2017late]. In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula. In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017]. In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk. This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.
